---
title: "Data Wrangling"
author: "Larry Alade"
date: "October 22, 2018"
output: html_document
---

```{r}
library(tidyverse) # install.packages("tidyverse")
```
## load and inspect dataset
```{r}
## read gapminder csv. Note the readr:: prefix identifies which package it's in
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
head(gapminder) # GIVES YOU THE FIRST SIX RECORDS BY DEFAULT
View(gapminder)
tail(gapminder) # list bottom six records by default
head(gapminder,3) # define number of records
str(gapminder)
summary(gapminder)


#Sidebar another way to look at summary data at a deeped level
install.packages("skimr")
skimr::skim(gapminder)

names(gapminder) #column names

dim(gapminder)
ncol(gapminder)
nrow(gapminder)

#reverse-engineer dim
c(nrow(gapminder), ncol (gapminder))


```

## Dplyr
```{r}
# filter by rows
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Mexico")
filter(gapminder, country %in% c("Mexico","Peru", "Brazil"))
filter(gapminder, country == "Mexico", year == 2002)

# find mean lifeExp of Sweden
sweden <- filter(gapminder, country == "Sweden")
names(sweden)
mean(sweden$lifeExp)

#select by columns
select(gapminder, year, lifeExp)
select(gapminder, -continent, -lifeExp)

#use select and filter together 
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)

```

## The pipe operator will change your life
```{r}
gapminder %>% head()
gapminder %>% head(3)

## instead of this...
gap_cambodia  <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp) 

## ...we can do this
gap_cambodia  <- gapminder %>% 
  filter(country == "Cambodia")%>% 
  select(-continent, -lifeExp) 
```


##Mutate adds new variables
```{r}
gapminder%>%
  mutate((index = 1:nrow(gapminder)))

gapminder %>%
  mutate(planet = "Earth")

gapminder %>%
## Challenge Find max GdP of Eqgypt and Vietnam.  create a new column (I want just one number that is max of those countries.)

gapminder %>%  
filter(country %in% c("Vietnam", "Egypt")) %>%
  mutate (gdp = pop * gdpPercap) %>%
  mutate(max_gdp = max(gdp))

gapminder %>%  
filter(country %in% c("Vietnam", "Egypt")) %>%
  mutate (gdp = pop * gdpPercap,
          max_gdp = max(gdp))

```

## group_by()
```{r}
## with mutate()
gap_grouped <- gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop * gdpPercap,
         max_gdp = max(gdp)) %>% 
ungroup()
View(gap_grouped)


## With sumarised ()
gap_sumarized <-gapminder %>% 
  group_by(country) %>% 
  mutate (gdp = pop*gdpPercap) %>% 
  sumarize(max_gdp = max(gdp)) %>% 
  ungroup() %>% 
View(gap_sumarized)



```

## Joins 
```{r}
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
View(co2)
names(co2)

co2 %>% head()
co2 %>% dim() # 12

## create new variable that is only 2007 data

gap_2007 <- gapminder %>%
  filter(year == 2007) 
gap_2007 %>% dim()

## left_join gap_2007 to co2
lj <- left_join(gap_2007, co2, by="country")
View(lj)

# another way to do this
lj <- gap_2007 %>% 
  left_join(co2, by = "country")

# the same as (note column sequence changes; error prone if analyses are based on column positioning)
lj2 <- left_join(co2, gap_2007, by = "country")

## right_join gap_2007 and co2
rj <- right_join(gap_2007, co2, by = "country")


```

## tidyverse

```{r}

library(tidyverse)
## gapminder data in a wide format
gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')

gap_wide %>% head()
View(gap_wide)
names(gap_wide)

```


##tidyr::gather()
```{r}
?gather
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country)

View(gap_long)
```


## tidyr::separate()
```{r}
gap_long <- gap_wide %>% 
  gather(key   = obstype_year,
         value = obs_values,
         -continent, -country) %>% 
separate(col = obstype_year,
         into = c("obs_type","year"),
         sep = "_", 
         convert = TRUE)
str(gap_long)
View(gap_long)
```

## plot long data
```{r}
canada_df <- gap_long %>% 
  filter(country == "Canada",
         obs_type == "lifeExp")

ggplot(data = canada_df, aes(x = year, y = obs_values))+ 
  geom_line() + 
  geom_smooth(span = 0.3)

```

plot the life expectancy of allcountries in the Americas

```{r}
life_df <- gap_long %>% 
  filter(obs_type =="lifeExp",
         continent == "Americas")

ggplot(data=life_df, aes(x = year, y=obs_values, color = country))+
  geom_line()

```

## Using gap_long, calculate and plot the the mean life expectancy for each continent over time from 1982 to 2007. Give your plot a title and assign x and y labels. Hint: use the dplyr::group_by() and dplyr::summarize() functions.

```{r}
continents <- gap_long %>% 
  filter(obs_type == "lifeExp", 
         year %in% 1980:2005) %>% 
  group_by(continent, year) %>% 
  summarize(mean_le = mean(obs_values))
View(continents)

ggplot(data = continents, aes(x = year, y = mean_le, color = continent)) + 
  geom_line() +
  labs(title = "Mean life expectancy",
       x = "Year",
       y = "Age (years)") 

## Additional customization
ggplot(data = continents, aes(x = year, y = mean_le, color = continent)) + 
  geom_line() +
  labs(title = "Mean life expectancy",
       x = "Year",
       y = "Age (years)",
       color = "Continent") +
  theme_classic() +
  scale_fill_brewer(palette = "Blues") 
  

```

## Exercise (Spread the data)

```{r}
head(gap_long) # remember the columns

gap_wide_new <- gap_long %>% 
  # first unite obs_type and year into a new column called var_names. Separate by _
  unite(col = var_names, obs_type, year, sep = "_") %>% 
  # then spread var_names out by key-value pair.
  spread(key = var_names, value = obs_values)

```
























































